---
title: "Initiation to R Software"
author: "Pierre Michel"
date: "Master AMSE 1st year, 2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem Set IV

## Exercice 1

a) Import the `data.frame` *iris* from the package `datasets`. The obtained `data.frame` contains the columns "Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width" and "Species". How many rows are there?
b) We focus on the distribution of the variable "Sepal.Length". Compute its mean (denoted $m$) and standard deviation (denoted $s$).
c) Study and comment the following graphs:

- histogram (`hist()`)
- tree (`stem()`)
- graph "Sepal.Length" versus normal distribution (`qqnorm()`)
- density (`plot(density())`)

Below is what you should get.

````{r, echo = F, fig.height = 4}
par(mfrow = c(1,3))
hist(iris$Sepal.Length, main = "Histogram")
stem(iris$Sepal.Length)
qqnorm(iris$Sepal.Length)
plot(density(iris$Sepal.Length), main = "Density plot")
```

d) Generate 8 samples of same size as "Sepal.Length", following a normal distribution $\mathcal{N}(m,s)$. Plot in the same graphical window: 9 histograms, 8 obtained through sample simulation, and the one of "Sepal.Length". Comment.
e) Do the same for the other graphs (`plot(density())` and `qqnorm()`).


## Exercice 2

a) Plot in a same graphical window 9 qq-plots, the first 3 are obtained with samples drawn from a normal distribution $\mathcal{N}(0,1)$ of size 10, the next 3 of size 100, the last 3 of size 1000. Comment.
b) Do the same thing for densities, with 9 samples drawn from a $\chi^2$ distribution; the first 3 with samples of size 100, the next 3 of size 1000, the last 3 of size 10000. Comment.

## Exercice 3

Import the dataset *airquality* from the package `cluster`. The `data.frame` obtained contains the columns "Ozone", "Solar.R", "Wind", "Temp", "Month", "Day". Plot the graph of "Ozone" in function of "Solar.R" (use `plot()`), then plot the same graph for the three levels of temperature ("Temp"), use `coplot()`. Comment. 
